Stage-Splitting a Modal Language

### 2014-10-15 paper/example.tex
added a few more notes about the example

To continue splitting the case statement, we need to us a sum type in the precomputation
to remind the code what branch its in.

@@

The final version is shown in figure...
Note that it's basically a tree algorithm...

The benefits of this decomposition are substantial: for a randomly ordered list of length $n$, 
the original \texttt{qSelect} takes expected $O(n)$ time per lookup, 
whereas \texttt{qSelect2} takes $O(\log n)$ time.
Since \texttt{qSelect1} does not depend on $k$, its $O(n\log n)$ cost can be amortized over many lookups.
%One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
%and then used to accelerate many future lookups for any value of $k$.

This paper defines how to statically perform splitting transformations, like the one described above, on programs written in a simple staged language we call \lang.  \lang\ unambiguously specifies how its terms should be split
% and features a type system that ensures valid splittings exist.
% only minimal refactoring should be required to write terms within this language,
%and from there, the type system should be enough to prove that a valid splitting exists.
by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.  
After splitting, those parts of the \lang\ term that were identified with stage \bbone\ 
will end up in the stage \bbone\ function ({\tt qSelect1}), 
and those parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function ({\tt qSelect2}).
Typing rules for \lang\ ensure a valid splitting exists.  That is, 
within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
but never from \bbtwo\ to \bbone.

>>>>

To continue splitting the case statement, we need to generate a sum type in the precomputation
to remind the code what branch its in.

@@

The final version, cleaned up and given better datatype names, is shown in \ref{fig:qssplit}.
Observe that \texttt{qSelect1} essentially builds a binary search tree from the input list, 
and \texttt{qSelect2} searches that tree for the correct index.
The benefits of this decomposition are substantial: for a randomly ordered list of length $n$, 
the original \texttt{qSelect} takes expected $O(n)$ time per lookup, 
whereas \texttt{qSelect2} takes $O(\log n)$ time.
Since \texttt{qSelect1} does not depend on $k$, its $O(n\log n)$ cost can be amortized over many lookups.

\nr{There's something else I need to say right here before the contribution statement below. 
I want the reader to be thinking "yeah, this splitting business does seem cool."
Bonus points if I also have them thinking "but it would be cooler if it were formal."}

This paper formally describes and analyzes the splitting transformation exemplified above.
(I also want to make it clear that recognizing \lang's appropriateness for this is itself is a contribution.)

We start by defining a simple staged language, called \lang, to express the input to our splitting algorithm.  
\lang\ unambiguously specifies how its terms should be split
% and features a type system that ensures valid splittings exist.
% only minimal refactoring should be required to write terms within this language,
%and from there, the type system should be enough to prove that a valid splitting exists.
by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.  
After splitting, those parts of the \lang\ term that were identified with stage \bbone\ 
will end up in the stage \bbone\ function ({\tt qSelect1}), 
and those parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function ({\tt qSelect2}).
Typing rules for \lang\ ensure a valid splitting exists.  That is, 
within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
but never from \bbtwo\ to \bbone.
